Cryptanalysis of \MORUS

### 2018-07-06 AC18_rebuttal.txt

Dear Tomer Ashur; Maria Eichlseder; Martin M. Lauridsen; Gaëtan Leurent; Brice Minaud; Yann Rotella; Yu Sasaki; Benoît Viguier,

Below please find the reviewer comments on your paper

  "Cryptanalysis of MORUS"

that was submitted to ASIACRYPT 2018.

https://secure.iacr.org/websubrev/ac2018/submit/rebuttal.php

Remember that the deadline for the rebuttals is July 13th, 11:45 UTC.

************************************************************************
* REBUTTAL *************************************************************
************************************************************************

We thank all reviewers for taking the time to review this submission.
We will take all editorial comments into account to prepare an updated version.
Below, we discuss some specific comments raised by the reviewers:

> A small drawback that I would add is that this paper would benefit from a
> concrete comparison with similar attacks, as the previous one that was found
> on AEGIS. While it does not low down the impact of the attack on a new scheme
> it could help to understand the differences for further cryptanalysis.

... TODO ...

> Moreover, the provided attack works either if the key is fixed or if the same
> plaintext is repeated a large number of times. While the former case is not a
> big limitation in practice even considering design recommendations, the later
> seems to be less probable in practice. Maybe examples of concrete scenarii
> should be emphasized to strengthen the proposal.

... TODO ...

************************************************************************
* REVIEWS **************************************************************
************************************************************************

This paper presents three attacks on MORUS: a linear attack on the
confidentiality of full MORUS, a
differential attack on the integrity of MORUS with reduced finalisation, and a
meet-in-the-middle
key-recovery attack on MORUS with reduced initialisation.

The main result, i.e. the linear attack, is quite interesting in itself, and
shows a potential
weakness in the parameters of MORUS-1280. In particular, the attack shows that
increasing the
internal state size alone from MORUS-640 to MORUS-1280 has little influence on
the linear
correlation. Additionally, it is very interesting that the correlation is
completely independent of
the key and nonce. While that attack requires repeated encryption of the same
plaintext with
different nonces, the authors argue that previous attacks on RC4 have been
carried out in this
setting.

Presentation wise, the authors introduce a small version of MORUS which, along
with excellent
figures, makes it significantly easier to grasp the details of the attack. It
is also nice to see
that the authors explicitly derive formulas for the linear expressions, and
additionally perform
experiments to back up these derivations.

For the two other attacks, while maybe not as interesting, the presentation is
equally clear and
easy to follow. All in all, the paper is well written as present valuable
cryptanalysis of a CAESAR
finalist. I recommend acceptance.

Some editorial comments:

- Page 6: Under finalization, tmp isn't defined. 
- The word "Symmetric" is a bit confusing. Maybe change to "Rotation
invariant"?
- It might be good to add the LAT for AND somewhere, as a help to the reader.

- Figure 3: Good, but needs a much more descriptive caption. Explain what the
numbers are, what the
x's are, annotate with round-numbers.
- I could use a bit more detail about the experimental verification in section
4.4.
- Page 21: "beginning and end of finalization" -> initialization
- Table 4 and 5 should be in the main body.

========================================================================

This paper presents new cryptanalysis against MORUS, one of the three
finalists in the CAESAR authenticated encryption competition in the group for
use-case high-speed software performance.

The main result of this paper is a keystream distinguisher on full MORUS-1280
based on linear approximations of its StateUpdate-function. 
It does not depend the secret key and thus does not allow key recovery, so in
principle it can only recover unknown bits of plaintext.
Against full MORUS-1280 with 256 bits of claimed security, one can recover
plaintext bits using about 2^152 encrypted blocks *under various keys*, which
is clearly impractical.
While MORUS claims 256-bit security and restricts each key to at most 2^64
encryptions, this does not hamper this attack.
Even though the data complexity of 2^152 is impractically large, as this is
the same scenario as the real-world decrypting plaintext attacks on RC4 in
TLS, this attack should be taken as a certificational break of full
MORUS-1280.

The linear approximation on the StateUpdate function can be seen as rotational
linear cryptanalysis, where every used linear mask on a 'register' (256-bit
for MORUS-1280) uses the same bits of each 'word' within a register (a
'register' consists of 4 'words').
The authors call this property 'symmetric', and this property is maintained
under various operations on linear masks: XORing two linear masks, performing
the same bit-wise rotation on each word in a register, and performing
word-wise rotations on a register.
This simplifies analysis and naturally gives rise to analysis where one just
considers the linear mask on the first word of each register (as the other
words have the same linear mask).
The authors introduce minimorus where each register just contains a single
word and remove all word-wise rotations.

The authors introduce a simplified analysis on simplified minimorus which
showcases their results and allowed experimental verification by the authors.
The full attack is then an extension of this analysis, where the authors
analyze the interaction of the non-linear components.

The paper is well written and provides new cryptanalysis, besides the main
result on full MORUS-1280 the paper includes attacks against round-reduce
initialization and finalization.

- consider renaming rotations similar to the existing terminology of 'bit-wise
rotation of machine registers'. i.e., X-wise rotation does rotation in steps
of unit X. This means that 'word-wise' rotation becomes 'bit-wise' rotation,
and 'register-wise' rotation becomes 'word-wise' rotation in this paper.
- consider renaming 'symmetric property' to 'rotational property' to make the
link to rotational differential cryptanalysis and to make it more specific
that it is a rotational symmetry instead of e.g. reflection symmetry.

========================================================================

Morus is a stream cipher based authenticated encryption scheme, typically
designed for high-performance applications. It is one of the finalist of the
CAESAR competition. As the major contribution of the paper, the authors have
presented a linear correlation in the keystream of full MORUS. For MORUS-1280
(-640), the authors have showed a correlation of $2^{-76}$ ($2^{-73}$ resp.),
which can be used to mount a distinguishing attack after $2^{152}$ ($2^{146}$
resp.) encryptions. This attack depicts that MORUS-1280-256 doesn't offer 
$256$-bit security for confidentiality, as claimed by the designers. The
authors have also presented one integrity attack and one key-recovery attacks
on reduced round MORUS.

The overall presentation of the paper is really good. The concept of using
symmetric linear combinations to deal with the bit rotations is novel. I also
appreciate the idea of building the linear trail for a reduced version (called
MiniMORUS) first, and then extending it to full MORUS using symmetrization.

The editorial quality of the paper is also good. Some minor issues of the
papers are:
1. abstract: “less than would be” -> less than what would be.
2. Same notation $C_i^t$ is used to denote a ciphertext as well as an equation
(page 11). 
3. “That the correlation ... 9 times” -> change the sentence.
4. Capitalization is not consistent (Keystream correlation, Data complexity,
Differential Trails, Full Diffusion Rounds etc).

Overall, the reported results of this paper are very interesting, especially
MORUS being one of the finalist of the ongoing CAESAR competition. Although
the reported attacks are not practical, it certainly breaks the claimed
security of MORUS-1280-256 and depicts some major weaknesses in the inherent
design of MORUS such as:
(i) It is possible to build trails irrespective of the rotation constants,
	(i)	(ii) Finalization phase involves some unnecessary non-linear computations
which are ultimately not used during the tag generation.
	(ii)	
With the light of the ongoing final round of the CAESAR competition, these
results are very relevant and might play a crucial role in the
selection/rejection of MORUS in the final portfolio; and hence I think the
work should be accepted.

========================================================================

This paper proposes several attacks on a family of authenticated encryption
schemes MORUS, which is one of finalists of the CAESAR competition. The
authors found linear correlations in the keystreams with 2^-73 and 2^-76 of on
MORUS-640 and MORUS-1280, respectively. These biases enable to develop
distinguishing attacks with 2^146 and 2^152 encryptions, respectively. Since
these linear correlations are independent from values of key, a typical
countermeasure such as key updating does not make sense, and broadcast setting
where the same plaintext is encrypted with different keys is directly applied.

Also, they analyze the reduced version of MORUS, and present forgery and key
recovery attacks on several reduced variants of MORUS.

The strongest result in this paper is the distinguishing attack on MORUS-1280,
because this attack successfully breaks the claimed security of 256-bit key
version of MORUS-1280 for the first time, while the required time and data
complexity is still very high and impractical.
Linear correlation attacks to stream ciphers are not new techniques but
finding good trails are non-trivial task. Furthermore, these attacks are
nicely backed up by experimental verifications of toy models. On the other
hands, the security analysis of the reduced variant of MORUS is less
interesting.

In the plaintext recovery attack in the broad setting, it requires more
samples than distinguishing attacks. So, the attack does not match the
practical attack scenario such as BEAST.

Overall, since MORUS is one of finalists of CAESAR, third party cryptanalysis
is valuable, I believe that the result should be published from this
conference

-The fact that correlations are independent from the key is not uncommon in
stream ciphers when the weakness of the key generation function is exploited,
e.g. RC4.

-In the plaintext recovery setting, it requires more samples than
distinguishing attacks. So, the attack does not match practical attack
scenario of BEAST.

========================================================================

This paper provides a cryptanalysis of the authenticated encryption MORUS,
finalist of the CAESAR competition. Its main contribution relies on a linear
correlation in the keystream of the full version of MORUS, which provides a
distinguisher on its output. The attack complexity for MORUS-1280 is lower
than what we would expect for a 256-bit security scheme. Two other smaller
contributions of this paper are forgery attack and key-recovery attack in the
nonce-misuse setting both in reduced versions of the authenticated encryption
scheme.

The editorial quality of this paper makes it very easy to follow. The authors
start with a concrete description of MORUS. Then they introduce a lighter
scheme they refer to as MiniMORUS to describe a simple version of their main
attack. Afterwards, the concrete attack of full MORUS is given, followed by
two extra contributions on round reduced versions of the scheme.

Strengths of this paper are listed below.

1) The main strength of this paper is the impact of the linear correlation
attack. Namely, the provided attack provides a way to distinguish MORUS'
output from random with a lower complexity than expected for such security.

2) The preliminary description of the attack on MiniMORUS is very appreciable
to understand the full attack. The trail fragments depicted in Figure 2 are
very helpful to get the procedure. Basically, the idea is to write equations
between ciphertext and registers in the state at different steps, using a zero
plaintext. Then, dependencies and equations' weights allow to conclude the
approximation.

3) The source code of MiniMORUS' attack is provided as a supplementary
material.

4) Going from MiniMORUS to full-MORUS is done through an elegant use of
symmetries. Furthermore, changing the zero plaintext to a known one is pretty
straightforward as the considered dependencies are linear.

5) The two extra attacks on initialization and finalization are not concrete
attacks in the sense that they only apply on reduced versions of MORUS.
Nevertheless, they nicely complete the analysis of MORUS.

A small drawback that I would add is that this paper would benefit from a
concrete comparison with similar attacks, as the previous one that was found
on AEGIS. While it does not low down the impact of the attack on a new scheme
it could help to understand the differences for further cryptanalysis.

Moreover, the provided attack works either if the key is fixed or if the same
plaintext is repeated a large number of times. While the former case is not a
big limitation in practice even considering design recommendations, the later
seems to be less probable in practice. Maybe examples of concrete scenarii
should be emphasized to strengthen the proposal. Nevertheless, as soon as it
is possible to mount this main attack under certain circumstances and because
this scheme is currently under evaluation, I think such work is mandatory to
determine the concrete security of proposals.

All in one, this paper presents huge contributions that totally fit the scope
of Asiacrypt conference. While I would appreciate more feedback on previous
related works, the paper is very well written and advantageously describes
every attack precisely enough for a 'non-symmetric' audience. For these
reasons, I lean towards acceptance.

Minor comments:
- few typos in README.nb

### 2018-07-10 AC18_rebuttal.txt

... TODO (Brice?)...

> Moreover, the provided attack works either if the key is fixed or if the same
> plaintext is repeated a large number of times. While the former case is not a
> big limitation in practice even considering design recommendations, the later
> seems to be less probable in practice. Maybe examples of concrete scenarii
> should be emphasized to strengthen the proposal.

... TODO ...

> In the plaintext recovery attack in the broad setting, it requires more
> samples than distinguishing attacks. So, the attack does not match the
> practical attack scenario such as BEAST.

  Brice wrote:
  "We are not aware of a distinguishing attack on MORUS with lower complexity than our plaintext recovery attack. Note that MORUS is not a block cipher, and distinguishing attacks on modes of operation (such as the CBC mode targeted in the original BEAST attack) do not apply."
  A small possibility is that in the broadcast setting of BEAST the same nonce would be used with different keys. If that's true the reviewer could be thinking of a generic multi-target key-guessing attack using the initial known plaintext segment, and they would be kind of right, except it's not a distinguishing attack.

  Gaëtan wrote:
  "Another option is that he might be talking about the small constant factors that are needed to get a high probability of success (with 1/epsilon^2 data we only have a small advantage)..  There is also a factor in the complexity depending on the number of bits/bytes that we want to recover (that we didn't consider in the paper, but are needed in order to recover a password or a web cookie)."

### 2018-07-10 AC18_rebuttal.txt

> A small drawback that I would add is that this paper would benefit from a
> concrete comparison with similar attacks, as the previous one that was found
> on AEGIS. While it does not low down the impact of the attack on a new scheme
> it could help to understand the differences for further cryptanalysis.

... TODO (Brice?)...

> Moreover, the provided attack works either if the key is fixed or if the same
> plaintext is repeated a large number of times. While the former case is not a
> big limitation in practice even considering design recommendations, the later
> seems to be less probable in practice. Maybe examples of concrete scenarii
> should be emphasized to strengthen the proposal.

As discussed in Section 6, repeated plaintexts actually appear in many practical
applications due to fixed formats (headers, metadata, file formats) and often
include some repeated secret parts (password, authentication cookie, ...).
This scenario has been exploited in several practical attacks, including BEAST
(TLS), RC4 attacks (TLS, WPA), and Sweet32 (TLS, OpenVPN).

@@

> In the plaintext recovery attack in the broad setting, it requires more
> samples than distinguishing attacks. So, the attack does not match the
> practical attack scenario such as BEAST.

Clearly, the complexity of the attack we propose is not practical, unlike the
attacks on legacy cryptosystems in other BEAST-style attacks.
However, the attack complexity is less than suggested by the security claim of
MORUS1280. We are not aware of a distinguishing attack on MORUS with lower
complexity than our plaintext recovery attack. Note that MORUS is not a block
cipher, and distinguishing attacks on modes of operation (such as the CBC mode
targeted in the original BEAST attack) do not apply.

  Brice wrote:
  A small possibility is that in the broadcast setting of BEAST the same nonce would be used with different keys. If that's true the reviewer could be thinking of a generic multi-target key-guessing attack using the initial known plaintext segment, and they would be kind of right, except it's not a distinguishing attack.

### 2018-07-10 morusAC_06_Discussion.tex
updated rebuttal

We emphasize that the correlation we uncover between plaintext and ciphertext bits is
\emph{absolute}, in the sense that it does not depend on the encryption
key, or on the nonce.  This is the same situation as the keystream correlations
in AEGIS~\cite{sacryptMinaud14}.  As such, they can be leveraged to
mount an attack in the broacast setting, where the same message is
encrypted multiple times with different IVs and potentially different
keys~\cite{DBLP:conf/fse/MantinS01}.  In particular, the broadcast setting appears
in practice in man-in-the-browser attacks against HTTPS connections
following the BEAST model~\cite{duong2011here}.  In this scenario, an
attacker uses Javascript code running in the victim's browser (by
tricking the victim to visit a malicious website) to generate a large
number of request to a secure website.  Because of details of the HTTP
protocol, each request includes an authentication token to identify the
user, and the attacker can target this token as a repeated plaintext.
Concretely, correlations in the RC4 keystream have been exploited in this
setting, leading to the recovery of authentication cookies in
practice~\cite{DBLP:conf/uss/AlFardanBPPS13}.

>>>>

We emphasize that the correlation we uncover between plaintext and ciphertext bits is
\emph{absolute}, in the sense that it does not depend on the encryption
key, or on the nonce.  This is the same situation as the keystream correlations
in AEGIS~\cite{sacryptMinaud14}.  As such, they can be leveraged to
mount an attack in the broadcast setting, where the same message is
encrypted multiple times with different IVs and potentially different
keys~\cite{DBLP:conf/fse/MantinS01}.  In particular, the broadcast setting appears
in practice in man-in-the-browser attacks against HTTPS connections
following the BEAST model~\cite{duong2011here}.  In this scenario, an
attacker uses Javascript code running in the victim's browser (by
tricking the victim to visit a malicious website) to generate a large
number of request to a secure website.  Because of details of the HTTP
protocol, each request includes an authentication token to identify the
user, and the attacker can target this token as a repeated plaintext.
Concretely, correlations in the RC4 keystream have been exploited in this
setting, leading to the recovery of authentication cookies in
practice~\cite{DBLP:conf/uss/AlFardanBPPS13}.
